Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

### 2021-08-25 2021/Entropy/entropy.tex
fix some mistakes in translation

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The article features the study of global optimization problems and numerical methods of their solution. Such tasks are computationally expensive since the objective function can be multi-extremal, nondifferentiable, and, as a rule, given in the form of a "black box". The study used a deterministic algorithm for finding the global extremum. The said algorithm is based neither on the concept of multistart, nor nature-inspired algorithms. The article provides computational rules of the one-dimensional algorithm and the nested optimization scheme which could be applied for solving complex multidimensional problems. Note that the solution complexity of global optimization problems essentially depends on the presence of multiple local extrema. In this paper, we apply machine learning methods to identify areas of attraction of local minima. The use of local optimization algorithms in the selected areas can significantly accelerate the convergence of global search as it could reduce the number of search trials in the vicinity of local minima. The results of computational experiments carried out on several hundred global optimization problems of different dimensionalities presented in the paper confirm the effect of accelerated convergence (in terms of the number of search trials required to solve a problem with a given accuracy).
}

@@

The successful application of machine learning (ML) methods to solve a wide range of problems leads to the emergence of new ways to apply ML for many tasks. 
Methods of machine learning were shown to be particularly effective for identifying the principal properties of the phenomena (for example, physical, economic, or social), which are stochastic by nature or contain some hidden parameters \cite{Golovenkin2020,Gonoskov2019}.
ML is also successfully used to solve complex problems of computational mathematics, for example, for simulation of dynamical systems  \cite{Seleznev2019}, solution of ordinary, partial, or stochastic differential equations \cite{Lagaris1998,Blechschmidt2021,Xu2020} .

For example, ML could be applied for solving such a complex problem of computational mathematics as global optimization.
The solution to this class of problems, as a rule, cannnot be found analytically and, therefore, one needs to construct numerical methods to solve it.

The numerical solution of optimization problems is fraught with significant difficulties. In many ways, they are related to the dimensionality and type of the objective function. Consequently, the most difficult tasks are those in which the objective function is multi-extremal, nondifferentiable, and, moreover, given in the form of a black box (i.e., in the form of some computational procedure, the input of which is an argument, and the output is the corresponding value of the function). These complex problems are the main focus of this article.

There are several approaches to the construction of numerical methods for solving global optimization problems. 
%Надо ли добавить алгоритмы на основе метамоделей ?
Some algorithms are based on the idea of a multistart: launching a local search either from different starting points or with varying parameters. Local optimization methods have a high convergence rate. At the same time, one of the main problems in multistart schemes is the choice of starting points that would correspond to the regions of attraction of various local solutions. 
Machine learning methods can be successfully applied to solve this problem.  
For example, in \cite{RinnooyKan1987} methods of cluster analysis were used to select promising starting points
In \cite{Cassioli2012} ], the area for launching the local method was allocated based on the classification of starting points using a support vector machine.

@@

Note that the algorithms of the last two classes do not provide guaranteed convergence to the solution of the problem and are inferior to deterministic algorithms in terms of the quality of solution \cite{Kvasov2018,Sergeyev2018} ] (eg, measured by the number of correctly solved problems from a particular set). Therefore, deterministic methods seem to be potentially more effective.

@@

Several strategies have been proposed to speed up the operation of the global search algorithm (in terms of the number of iterations required to solve the problem with a given accuracy). In this paper, we propose a new approach to acceleration based on identifying areas of attraction of local minima using machine learning methods. The identification of regions of attraction and the launch of local search in these regions can significantly reduce the number of trials required for the method to achieve global convergence. Experiments carried out on a series of several hundred test problems confirm this statement.

@@

In this paper, we will consider global optimization problems of the form
\begin{eqnarray}\label{main_problem}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in \text{R}^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}
Problem (\ref{main_problem}) is considered under the assumption that the objective function is multi-extreme, is given in a form of a "black box" , and the calculation of its values is associated with solving the problem of numerical modeling, which makes the solution a labor-intensive operation.

A typical situation for many applied problems is when a limited change in the vector of parameters y causes a limited change in the values of $\varphi(y)$. The mathematical model describing this premise is based on the assumption that the Lipschitz condition is satisfied
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty.
\]
The common assumption for many approaches to the development of optimization algorithms is that the objective function is Lipschitzian  \cite{Jones1993,Pinter1996,Zilinskas2008,Evtushenko2009}.
At the same time, many known approaches are based on various methods of dividing the search domain into a system of subdomains and then choosing the most promising subdomain for placing the next trial (calculating the value of the objective function)  \cite{Jones2009,Zilinskas2010,Evtushenko2013,Kvasov2013,Paulavicius2016}. 
An important property of global optimization problems is the fact that, in contrast to the problems of finding a local extremum, the global minimum is an integral characteristic of the problem being solved. Making sure that the point  $y^*\in D$ is a solution to the problem requires going beyond its neighborhood to the investigation of the entire search domain. As a result, when minimizing substantially multi-extremal functions, the numerical method must construct a coverage of the search domain. As a result, the number of nodes of this coverage increases exponentially with increasing dimensionality. 
This feature determines the high complexity of solving multiextremal optimization problems making dimensionality a critical factor that affects the complexity of their solution.

The dimensionality in multi-extreme optimization leads to many problems, so, scholars use a wide variety of approaches to reducing it.  
For example, simplicial or diagonal partition of the search domain allows using methods for solving one-dimensional problems to solve the original multidimensional problem (see, for example,  \cite{PaulaviciusZilinskas2014,Sergeyev2017,Sergeyev2013}). 
Another well-known approach to dimensionality reduction is using the Peano space-filling curves to map the multidimensional domain onto a one-dimensional interval \cite{Strongin2000}.

In this work, we will use another method based on the nested optimization scheme \cite{Shi2000,Grishagin2001,VanDam2010,Grishagin2015} and its generalization \cite{Grishagin2016,Grishagin2016_1}.
The nested optimization scheme, on the one hand, does not worsen the properties of the minimized function (unlike reduction using Peano curves), and, on the other hand, does not require the use of complex data structures to support simplex or diagonal partitions of the feasible region
At the same time, the nested optimization scheme makes it possible to reduce the solution of a multidimensional problem to the solution of a set of standard one-dimensional problems, which, then, can be solved by a wide range of one-dimensional global optimization algorithms.

@@

As a standard, let us consider a one-dimensional multiextremal optimization problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
with an objective function satisfying the Lipschitz condition

Here is the description of global search algorithm (GSA) for solving the basic problem in accordance with \cite{Strongin2000}.
]. In the course of its work, GSA generates a sequence of points  $x^i$, at which the values of the objective function  $z^i=\varphi(x^i)$ are calculated. 
We will refer to the process of calculating the value of the objective function as  \textit{trial}.

In accordance with the algorithm, the first two trials are carried out at the boundary points of the segment $[0,1]$, т.е. $x^0=0,\;x^1=1$. 
At these points, the values of the objective function  $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ и уare calculated and the counter value is set to $k=1$. 
The point of the next trial $x^{k+1}, k\geq 1,$ is selected in accordance with the following procedure:

 Step 1. Reindex  (starting at 0) the points xl, $x^i,\:0\leq i\leq k$, of the trials conducted  in ascending order of the coordinate, i.e.
\begin{equation}\label{xt}
0=x_0<x_1<\ldots <x_{k}=1.
\end{equation} 
Compare the values of the objective function $z_i=\varphi(x_i), \; 0\leq i\leq k$ to the points $x_i, \; 0\leq i\leq k$, at which these values were calculated.

Step 2. Calculate the maximum absolute value of the relative first difference
\begin{equation}\label{mu}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i},
\end{equation}
where $\Delta_i = x_i-x_{i-1}$. If the value calculated in accordance with this formula is equal to zero, then take $\mu = 1$.

@@

Step 4. Find interval $(x_{t-1},x_t)$ with the maximum characteristics
\begin{equation}\label{MaxR}
R(t)=\max_{1\leq i\leq {k}}R(i).
\end{equation}
If the maximum characteristics corresponds to several intervals, then choose the minimum number that satisfies as $t$ (\ref{MaxR}).

@@

The algorithm stops when the condition  $\Delta_t<\epsilon$ is satisfied; here $t$ from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy. 
The following values are selected to evaluate the efficiency of the solution
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i).
\]

The theoretical conditions that determine the convergence of the algorithm are presented in \cite{Strongin2000}. 
The operation of the algorithm during the minimization of a specific multiextremal function, which is specified in accordance with formula  (\ref{hill}), is shown in Figure \ref{fig1}. The algorithm was launched with the parameter $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$in the criterion for stopping the method.
Figure \ref{fig1} shows the objective function graph and points of 71 search trials which GSA needed to solve the problem to the specified accuracy. This highlights the problem of all methods of global optimization - the concentration of points in the vicinity of local minima of the problem, which are not a global solution.

@@

\subsection{Machine Learning Regression as a tool for identifying areas of attraction of local extrema}\label{TreeGSA}

The functions considered in this study belong to the class of Lipschitzian functions. Therefore, classical regression methods (for example, polynomial regression, where a function is approximated by a polynomial of a given degree) will not properly match the behavior of the function.
A more powerful tool for this task is regression splines. When constructing a regression spline, the domain of definition is divided into K non-overlapping subdomains. In each of such domains, the function is approximated by a polynomial. Dividing the interval into a sufficient number of subdomains allows one to very accurately approximate the original function.

@@

Therefore, within the framework of the study, we chose a regression model based on decision trees to analyze the local behavior of a function.
For example, if the objective function is properly approximated by polynomials, then the polynomial regression, of course, will appropriately convey the properties of the function. However, if there is a more complex relationship between them, then the decision tree can surpass the classical variants of regression in terms of the quality of the approximation. 
%Здесь нужен рисунок - функция Шекеля, и два варианта регрессии - полином и деревья решений. Шекель должен плохо полиномами представляться, т.к. там функции вида 1/x
At the same time, the regression based on the decision tree makes it possible to easily identify the areas of attraction of local extrema with sufficient accuracy.

Building a regression using a decision tree consists of two main steps:
\begin{itemize}
	\item Search domain $D$ is divided into $J$ non-overlapping subdomains  $D_1, D_2, ..., D_J$, provided that $D = \bigcup_{j=1}^{J}{D_j}$.
	\item Any value falling into the subdomain $D_j$, i.e. $x \in D_j$, is matched to the average value  $c_j$ based on the training trial values that fall into this subdomain. 
\end{itemize}

@@

Let $x^{k+1}$ be the point of the current trial. For a given point, an index  $j$ is sought such that $x^{k+1} \in D_j$. 
Next, the  $c_j$ values corresponding to neighboring subdomains are compared. If one of the conditions is met
\begin{gather}\label{LocalPointC}
c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3} \leq c_{j+4}, \; j=1; \nonumber \\ 
c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3}, \; j=2;  \nonumber \\  
c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2}, \; 3 \leq j \leq J-2; \\ 
c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1}, \; j=J-1; \nonumber \\ 
c_{j-4} \geq c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j}, \; j=J; \nonumber 
\end{gather}
then the subdomain $D_j$ is considered the area of attraction of the local minimum. Here, one can start a local search, and subsequently exclude this subdomain from the global search. 
Any of the zero-order local methods can be used as a local algorithm, for example, golden section search or parabolic interpolation \cite{Press}.

In order to modify the global search algorithm from subsection  \ref{CoreGSA} to exclude the regions of attraction of local minima, we associate each test point  $x^i$, obtained during the operation of the algorithm with an additional criterion $q^i \in \{0,1,2\}$, which will characterize the properties of this point. 
The value  $q^i=0$ is assigned by default and indicates that point $x^i$ is obtained as a result of rule (\ref{xk1}) of the global search algorithm.
The value $q^i=1$ is assigned to the points obtained as a result of the work of the local method, while the value  $q^i=2$ corresponds to the point of the local minimum found as a result of the work of the local search.

@@

Steps 1-5 of the GSA-DT algorithm are the same as steps 1-5.

@@

Step 7. Construct a decision tree based on the results of the tests performed, and obtain the corresponding piecewise linear approximation, which assigns the value $c_1, c_2, ..., c_J$ to each subdomain  $D_1, D_2, ..., D_J$.

Step 8. For the point $x^{k+1}$ of the current trial, find a number $j$ such that $x^{k+1} \in D_j$ and check whether the condition  (\ref{LocalPointC}) is satisfied.If condition (\ref{LocalPointC}) is satisfied, start a local search in the domain $D_j$ from the point  $x^{k+1}$. 
The results of all trials performed during the local search at the points $x^{k+2}, ...,x^{k+k_{local}}$ are stored in the information base of the algorithm and are used at subsequent iterations. 
All these points receive the attribute  $q^i=1$, $i = k+2, ... , k+k_{local}$. The attribute equal to 2 is assigned to the point corresponding to the found local minimum.

@@

\begin{itemize}
	\item $|x_{t} - x_{t-1}|<\epsilon$, where $t$ from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy;
	\item $|x^{k+1} - x_{t-1}|<\epsilon$ and $q_{t-1} = 2$,
	\item $|x^{k+1}-x_{t}|<\epsilon$ and $q_{t} = 2$.
\end{itemize}

For example, consider the operation of the GSA-DT algorithm with minimization of the same multi-extreme function, which is presented  Figure \ref{fig1}. The same parameters were used at the start of the algorithm: the parameter  $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$ in the stopping criterion of the method. 
Figure \ref{fig2} illustrates the operation of the GSA-DT algorithm. In addition to the graph of the objective function, it shows a piecewise constant approximation of the form (\ref{dtree}) built at the final stage of the search. Black points on the graph correspond to the global search phase, green points correspond to the work of the local method. In total, the GSA-DT method required 49 trials to solve the problem; there was no accumulation of trial points in the vicinity of local minima.

@@

Then, in accordance with  (\ref{nested}), solving the multidimensional problem  (\ref{main_problem}) is reduced to solving a one-dimensional problem 
\begin{equation}\label{nested_1}
\varphi^* = \min_{y_1\in\left[a_1,b_1\right]}\varphi_1(y_1).
\end{equation}
However, each calculation of the value of the function $y_1$ at some fixed point y1 presupposes the solution of the one-dimensional optimization problem of the second level
\begin{equation}
\varphi_1(y_1) = \min_{y_2\in\left[a_2,b_2\right]}\varphi_2(y_1,y_2).
\end{equation}
Calculation of the values of the function  $\varphi_2$ in turn, requires one-dimensional minimization of the function  $\varphi_3$ all the way to the solution of the problem
\begin{equation}
\varphi_{N-1}(y_1,...,y_{N-1}) = \min_{ y_{N} \in\left[a_{N},b_{N}\right]} \varphi_{N}(y_1,...,y_{N})
\end{equation}
at the last level of recursion.

The solution of the set of subtasks arising in the nested optimization scheme  (\ref{nested_i}) can be organized in different ways. 
The obvious way (described in detail in \cite{Grishagin2001,Grishagin2015}) is based on solving subproblems in accordance with the recursive order of their generation. However, a significant part of the information about the objective function is lost here.

@@

Note that within the framework of the original nested optimization scheme, the generated subproblems are solved only sequentially; the resulting hierarchical scheme for generating and solving subproblems has the form of a tree. The construction of this tree occurs dynamically in the process of solving the original problem (\ref{main_problem}). In this case, the calculation of one value of the function $\varphi_i(y_1,y_2,...,y_i)$ at the $i$-th ith level requires a complete solution of all problems of one of the subtrees of level  $i+1$.

@@

Let nested subproblems of the form (\ref{nested_i}) be solved using the global search algorithm described in subsection  \ref{CoreGSA}. Then each subproblem  (\ref{nested_i})  can be assigned some numerical value called a characteristic of this problem. As such a characteristic, we can take the value $R(t)$ from (\ref{MaxR}), i.e. the maximum characteristic from the characteristics of the intervals formed in this task. The higher the values of this characteristic, the more promising is the search for the global minimum of the original problem in the given subproblem (\ref{main_problem}).Therefore, at each iteration, a subproblem with the maximum characteristic is selected, and a trial is carried out in it. This either leads to the calculation of the value of the objective function  $\varphi(y)$(if the selected subtask belonged to the level  $j=M$), or generates new subproblems according to  (\ref{nested_i}) for $j\leq M-1$. In the latter case, new generated tasks are added to the current set of tasks, their characteristics are calculated, and the process is repeated. The completion of the optimization process occurs when the condition for stopping the algorithm that solves this problem is satisfied for the root problem.

@@

Numerical experiments were performed on the Lobachevsky supercomputer of the University of Nizhny Novgorod (operating system CentOS 7.2, management system SLURM). One supercomputer node has 2 Intel Sandy Bridge E5-2660 2.2 GHz processors, 64 Gb RAM. The CPU is 8-core (i.e. a total of 16 CPU cores are available on the node). All the algorithms were implemented in Cd— +; GCC 5.5.0 was used for compilation on the supercomputer.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The paper features the study of global optimization problems and numerical methods of their solution. Such problems are computationally expensive since the objective function can be multi-extremal, nondifferentiable, and, as a rule, given in the form of a ``black box''. The study used a deterministic algorithm for finding the global extremum. The said algorithm is based neither on the concept of multistart, nor nature-inspired algorithms. The article provides computational rules of the one-dimensional algorithm and the nested optimization scheme which could be applied for solving multidimensional problems. Note that the solution complexity of global optimization problems essentially depends on the presence of multiple local extrema. In this paper, we apply machine learning methods to identify regions of attraction of local minima. The use of local optimization algorithms in the selected regions can significantly accelerate the convergence of global search as it could reduce the number of search trials in the vicinity of local minima. The results of computational experiments carried out on several hundred global optimization problems of different dimensionalities presented in the paper confirm the effect of accelerated convergence (in terms of the number of search trials required to solve a problem with a given accuracy).
}

@@

The successful application of machine learning (ML) methods to solve a wide range of problems leads to the emergence of new ways to apply ML for many tasks. 
Methods of machine learning were shown to be particularly effective for identifying the principal properties of the phenomena (for example, physical, economic, or social), which are stochastic by nature or contain some hidden parameters \cite{Golovenkin2020,Gonoskov2019}.
ML is also successfully used to solve complex problems of computational mathematics, for example, for simulation of dynamical systems  \cite{Seleznev2019}, solution of ordinary, partial, or stochastic differential equations \cite{Lagaris1998,Blechschmidt2021,Xu2020}.

For example, ML could be applied for solving such a complex problem of computational mathematics as global optimization.
The solution to this class of problems, as a rule, cannot be found analytically and, therefore, one needs to construct numerical methods to solve it.

The numerical solution of optimization problems is fraught with significant difficulties. In many ways, they are related to the dimensionality and type of the objective function. Consequently, the most difficult problems are those in which the objective function is multi-extremal, nondifferentiable, and, moreover, given in the form of a ``black box'' (i.e., in the form of some computational procedure, the input of which is an argument, and the output is the corresponding value of the function). These complex problems are the main focus of this article.

There are several approaches to the construction of numerical methods for solving global optimization problems. 
%Надо ли добавить алгоритмы на основе метамоделей ?
Some algorithms are based on the idea of a multistart: launching a local search either from different starting points or with varying parameters. Local optimization methods have a high convergence rate. At the same time, one of the main problems in multistart schemes is the choice of starting points that would correspond to the regions of attraction of various local solutions. 
Machine learning methods can be successfully applied to solve this problem.  
For example, in \cite{RinnooyKan1987} methods of cluster analysis were used to select promising starting points.
In \cite{Cassioli2012}, the area for launching the local method was allocated based on the classification of starting points using a support vector machine.

@@

Note that the algorithms of the last two classes do not provide guaranteed convergence to the solution of the problem and are inferior to deterministic algorithms in terms of the quality of solution \cite{Kvasov2018,Sergeyev2018} (e.g., measured by the number of correctly solved problems from a particular set). Therefore, deterministic methods seem to be potentially more effective.

@@

Several strategies have been proposed to speedup the global search algorithm (in terms of the number of iterations required to solve the problem with a given accuracy). In this paper, we propose a new approach to acceleration based on identifying areas of attraction of local minima using machine learning methods. The identification of regions of attraction and the launch of local search in these regions can significantly reduce the number of trials required for the method to achieve global convergence. Experiments carried out on a series of several hundred test problems confirm this statement.

@@

In this paper, we will consider global optimization problems of the form
\begin{eqnarray}\label{main_problem}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in \text{R}^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}
Problem (\ref{main_problem}) is considered under the assumption that the objective function is multi-extreme, is given in a form of a ``black box'', and the calculation of its values is associated with solving the problem of numerical simulation, which makes the solution a labor-intensive operation.

A typical situation for many applied problems is when a limited change in the vector of parameters $y$ causes a limited change in the values of $\varphi(y)$. The mathematical model describing this premise is based on the assumption that the Lipschitz condition is satisfied
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty.
\]
This assumption is typical for many approaches to the development of optimization algorithms \cite{Jones1993,Pinter1996,Zilinskas2008,Evtushenko2009}.
At the same time, many known approaches are based on various methods of dividing the search domain into a system of subdomains and then choosing the most promising subdomain for placing the next trial (calculating the value of the objective function)  \cite{Jones2009,Zilinskas2010,Evtushenko2013,Kvasov2013,Paulavicius2016}. 
An important property of global optimization problems is the fact that, in contrast to the problems of finding a local extremum, the global minimum is an integral characteristic of the problem being solved. Making sure that the point  $y^*\in D$ is a solution to the problem requires going beyond its neighborhood to the investigation of the entire search domain. As a result, when minimizing substantially multi-extremal functions, the numerical method must construct a coverage of the search domain. The number of nodes of this coverage increases exponentially with increasing dimensionality. 
This feature determines the high complexity of solving multiextremal optimization problems making dimensionality a critical factor that affects the complexity of their solving.

The dimensionality in multi-extremal optimization leads to many issues, so, scholars use a wide variety of approaches to reducing it.  
For example, simplicial or diagonal partition of the search domain allows using methods for solving one-dimensional problems to solve the original multidimensional problem (see, for example,  \cite{PaulaviciusZilinskas2014,Sergeyev2017}). 
Another well-known approach to dimensionality reduction is using the Peano space-filling curves to map the multidimensional domain onto a one-dimensional interval \cite{Strongin2000,Sergeyev2013}.

In this work, we will use another method based on the nested optimization scheme \cite{Shi2000,Grishagin2001,VanDam2010,Grishagin2015} and its generalization \cite{Grishagin2016,Grishagin2016_1}.
The nested optimization scheme, on the one hand, does not worsen the properties of the objective function (unlike reduction using Peano curves), and, on the other hand, does not require the use of complex data structures to support simplex or diagonal partitions of the feasible region.
At the same time, the nested optimization scheme makes it possible to reduce the original multidimensional optimization problem to a family of recursively connected one-dimensional optimization subproblems, which can be solved by a wide range of one-dimensional global optimization algorithms.

@@

As a standard, let us consider a one-dimensional multiextremal optimization problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
with an objective function satisfying the Lipschitz condition.

Here is the description of global search algorithm (GSA) for solving the basic problem in accordance with \cite{Strongin2000}.
In the course of its work, GSA generates a sequence of points  $x^i$, at which the values of the objective function  $z^i=\varphi(x^i)$ are calculated. 
We will refer to the process of calculating the value of the objective function as \textit{trial}.

In accordance with the algorithm, the first two trials are carried out at the boundary points of the segment $[0,1]$, т.е. $x^0=0,\;x^1=1$. 
At these points, the values of the objective function  $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ are calculated and the counter value is set to $k=1$. 
The point of the next trial $x^{k+1}, k\geq 1,$ is selected in accordance with the following procedure.

 Step 1. Renumber  (starting at 0) the points $x^i,\:0\leq i\leq k$, of the trials conducted  in ascending order of the coordinate, i.e.
\begin{equation}\label{xt}
0=x_0<x_1<\ldots <x_{k}=1.
\end{equation} 
Associate the values of the objective function $z_i=\varphi(x_i), \; 0\leq i\leq k,$ to the points $x_i, \; 0\leq i\leq k$, at which these values were calculated.

Step 2. Calculate the maximum absolute value of the relative first difference
\begin{equation}\label{mu}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i},
\end{equation}
where $\Delta_i = x_i-x_{i-1}$. If the value calculated in accordance with (\ref{mu}) is equal to zero, then take $\mu = 1$.

@@

Step 4. Find the interval $(x_{t-1},x_t)$ with the maximum characteristic
\begin{equation}\label{MaxR}
R(t)=\max_{1\leq i\leq {k}}R(i).
\end{equation}
If the maximum characteristic corresponds to several intervals, then choose the minimum number that satisfies (\ref{MaxR}) as $t$.

@@

The algorithm stops when the condition  $\Delta_t<\epsilon$ is satisfied; here $t$ is from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy. 
The values
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i)
\] 
are selected to estimate the solution

The theoretical conditions that determine the convergence of the algorithm are presented in \cite{Strongin2000}. 
The work of the algorithm during the minimization of a specific multiextremal function, which is specified in accordance with formula  (\ref{hill}), is shown in Figure \ref{fig1}. The algorithm was launched with the parameter $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$ in the method stopping condition.
Figure \ref{fig1} shows the objective function graph and points of 71 search trials which GSA needed to solve the problem to the specified accuracy. This highlights the problem of all methods of global optimization --- the concentration of trial points in the vicinity of local minima of the problem, which are not a global solution.

@@

\subsection{Machine Learning Regression as a tool for identifying attraction regions of local extrema}\label{TreeGSA}

The functions considered in this study belong to the class of Lipschitzian functions. Therefore, classical regression methods (for example, polynomial regression, where a function is approximated by a polynomial of a given degree) will not properly match the behavior of the function.
A more powerful tool for this task is regression splines. When constructing a regression spline, the domain is divided into $K$ non-overlapping subdomains. In each of such domains, the function is approximated by a polynomial. Dividing the interval into a sufficient number of subdomains allows one to very accurately approximate the original function.

@@

Therefore, within the framework of the study, we chose a regression model based on decision trees to analyze the local behavior of a function.
For example, if the objective function is properly approximated by polynomials, then the polynomial regression, of course, will appropriately convey the properties of the function. However, if there is a more complex relationship between them, then the decision tree can surpass the classical variants of regression in terms of the quality of the approximation. 
At the same time, the regression based on the decision tree makes it possible to easily identify the areas of attraction of local extrema with sufficient accuracy.

Building a regression using a decision tree consists of two main steps:
\begin{itemize}
	\item Search domain $D$ is divided into $J$ non-overlapping subdomains  $D_1, D_2, ..., D_J$, provided that $D = \bigcup_{j=1}^{J}{D_j}$.
	\item Any value falling into the subdomain $D_j$, i.e. $x \in D_j$, is matched to the average value  $c_j$ based on the training trials that fall into this subdomain. 
\end{itemize}

@@

Let $x^{k+1}$ be the point of the current trial. For a given point, an index  $j$ is sought such that $x^{k+1} \in D_j$. 
Next, the  $c_j$ values corresponding to neighboring subdomains are compared. If one of the conditions is met
\begin{gather}
c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3} \leq c_{j+4}, \; j=1; \nonumber \\ 
c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3}, \; j=2;  \nonumber \\  
c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2}, \; 3 \leq j \leq J-2; \label{LPC} \\ 
c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1}, \; j=J-1; \nonumber \\ 
c_{j-4} \geq c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j}, \; j=J; \nonumber 
\end{gather}
then the subdomain $D_j$ is considered the area of attraction of the local minimum. Here, one can start a local search, and subsequently exclude this subdomain from the global search. 
Any of the zero-order local methods can be used as a local algorithm, for example, golden section search or parabolic interpolation \cite{Press}.

In order to modify the global search algorithm from subsection  \ref{CoreGSA} to exclude the regions of attraction of local minima, we associate each trial point  $x^i$ obtained during the operation of the algorithm with an additional attribute $q^i \in \{0,1,2\}$, which will characterize the properties of this point. 
The value  $q^i=0$ is assigned by default and indicates that point $x^i$ is obtained as a result of rule (\ref{xk1}) of the global search algorithm.
The value $q^i=1$ is assigned to the points obtained as a result of the work of the local method, while the value  $q^i=2$ corresponds to the point of the local minimum found as a result of the work of the local search.

@@

Steps 1-5 of the GSA-DT algorithm are the same as steps 1-5 of GSA.

@@

Step 7. Construct a decision tree based on the results of the trials performed, and obtain the corresponding piecewise constant approximation, which assigns the value $c_1, c_2, ..., c_J$ to each subdomain  $D_1, D_2, ..., D_J$.

Step 8. For the point $x^{k+1}$ of the current trial, find a number $j$ such that $x^{k+1} \in D_j$ and check whether the condition  (\ref{LPC}) is satisfied. If condition (\ref{LPC}) is satisfied, start a local search in the domain $D_j$ from the point  $x^{k+1}$. 
The results of all trials performed during the local search at the points $x^{k+2}, ...,x^{k+k_{local}}$ are stored in the information base of the algorithm and are used at subsequent iterations. 
All these points receive the attribute  $q^i=1$, $i = k+2, ... , k+k_{local}$. The attribute equal to 2 is assigned to the point corresponding to the found local minimum.

@@

\begin{itemize}
	\item $|x_{t} - x_{t-1}|<\epsilon$;
	\item $|x^{k+1} - x_{t-1}|<\epsilon$ and $q_{t-1} = 2$;
	\item $|x^{k+1}-x_{t}|<\epsilon$ and $q_{t} = 2$;
\end{itemize}
where $t$ is from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy.

For example, consider the work of the GSA-DT algorithm with minimization of the same multi-extremal function, which is presented in Figure \ref{fig1}. The same parameters were used at the start of the algorithm: the parameter  $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$ in the stopping criterion of the method. 
Figure \ref{fig2} illustrates the operation of the GSA-DT algorithm. In addition to the graph of the objective function, it shows a piecewise constant approximation of the form (\ref{dtree}) built at the final stage of the search. Black points on the graph correspond to the global search phase, green points correspond to the work of the local method. In total, the GSA-DT method required 49 trials to solve the problem; there was no accumulation of trial points in the vicinity of local minima.

@@

Then, in accordance with  (\ref{nested}), solving the multidimensional problem  (\ref{main_problem}) is reduced to solving a one-dimensional problem 
\begin{equation}\label{nested_1}
\varphi^* = \min_{y_1\in\left[a_1,b_1\right]}\varphi_1(y_1).
\end{equation}
However, each calculation of the value of the function $\varphi_1$ at some fixed point $y_1$ presupposes the solution of the one-dimensional optimization problem of the second level
\begin{equation}
\varphi_1(y_1) = \min_{y_2\in\left[a_2,b_2\right]}\varphi_2(y_1,y_2).
\end{equation}
Calculation of the values of the function  $\varphi_2$ in turn, requires one-dimensional minimization of the function  $\varphi_3$ all the way to the solution of the problem
\begin{equation}
\varphi_{N-1}(y_1,...,y_{N-1}) = \min_{ y_{N} \in\left[a_{N},b_{N}\right]} \varphi_{N}(y_1,...,y_{N})
\end{equation}
at the last level of recursion.

The solution of the set of subproblems arising in the nested optimization scheme  (\ref{nested_i}) can be organized in different ways. 
The obvious way (described in detail in \cite{Grishagin2001,Grishagin2015}) is based on solving subproblems in accordance with the recursive order of their generation. However, a significant part of the information about the objective function is lost here.

@@

Note that within the framework of the original nested optimization scheme, the generated subproblems are solved only sequentially; the resulting hierarchical scheme for generating and solving subproblems has the form of a tree. The construction of this tree occurs dynamically in the process of solving the original problem (\ref{main_problem}). In this case, the calculation of one value of the function $\varphi_i(y_1,y_2,...,y_i)$ at the $i$-th level requires a complete solution of all problems of one of the subtrees of level  $i+1$.

@@

Let nested subproblems of the form (\ref{nested_i}) be solved using the global search algorithm described in subsection  \ref{CoreGSA}. Then each subproblem  (\ref{nested_i})  can be assigned some numerical value called a characteristic of this problem. As such a characteristic, we can take the value $R(t)$ from (\ref{MaxR}), i.e. the maximum characteristic from the characteristics of the intervals formed in this task. The higher the values of this characteristic, the more promising is the search for the global minimum of the original problem (\ref{main_problem}) in the given subproblem.Therefore, at each iteration, a subproblem with the maximum characteristic is selected, and a trial is carried out in it. This either leads to the calculation of the value of the objective function  $\varphi(y)$ (if the selected subtask belonged to the level  $j=M$), or generates new subproblems according to (\ref{nested_i}) for $j\leq M-1$. In the latter case, new generated tasks are added to the current set of tasks, their characteristics are calculated, and the process is repeated. The completion of the optimization process occurs when the condition for stopping the algorithm that solves this problem is satisfied for the root problem.

@@

Numerical experiments were performed on the Lobachevsky supercomputer of the University of Nizhny Novgorod (operating system CentOS 7.2, management system SLURM). One supercomputer node has 2 Intel Sandy Bridge E5-2660 2.2 GHz processors, 64 Gb RAM. The CPU is 8-core (i.e. a total of 16 CPU cores are available on the node). All the algorithms were implemented in C++; GCC 5.5.0 was used for compilation on the supercomputer.
